---
title: "EOW_camtrapdp_tutorial"
author: "Martijn Bollen"
date: "2025-05-05"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# 0. Introduction

In this tutorial I will show how to import, filter and pre-process camera trap data using the `camtrapdp` package.
By following the step-by-step process in this tutorial, you will be able to share data in the context of the [EOW](https://www.eowilsonfoundation.org) biodiversity monitoring using camera traps.

Let's start by loading the required packages

```{r setup, include=FALSE}
library(camtrapdp)
library(lubridate)
library(purrr)
library(tidyr)
library(dplyr)
library(rgbif)
```

For this tutorial we will rely on an example dataset, published on GBIF and accessed through the `rgbif` package.
Available datasets include:

```{r rgbif datasets}
gbif_datasets <- function() {
  datasets <- occ_count(facet = "datasetKey", protocol = "CAMTRAP_DP")
  safe_dataset_get <- possibly(dataset_get, otherwise = NULL)
  
  datasets <- datasets |>
    rowwise() |>
    mutate(metadata = list(safe_dataset_get(datasetKey))) |>
    unnest_wider(metadata)
  
  return(datasets)
}
  
datasets <- gbif_datasets()
datasets
```

Now let's download the "Dataset from Snapshot Japan 2023" as our example dataset.

```{r rgbif download}
dataset <- datasets |> filter(title == "Dataset from Snapshot Japan 2023")
ctdp_path <- tempdir()
zip_path <- file.path(ctdp_path, "snapshot_japan.zip")
download.file(url = dataset$endpoints[[1]][[1]]$url, zip_path, method = "curl")
```

# 1. Import Camtrap DP datasets

Now that we have downloaded our example dataset, it is time to import the data.
If you have exported your own dataset from Agouti, or another file manager, you should start your workflow here.

```{r import}
unzip(zip_path, exdir = tempdir())
ctdp <- read_camtrapdp(file.path(tempdir(), "datapackage.json"))
proj <- ctdp$name

# Check the structure of the data package
ctdp
```

# 2. Identify parsing problems, if any

Before filtering the dataset to the EOW survey period, we will check if there are any parsing problems in either the `deployments`, `media` or `observations` tables.

```{r parsing problems}
# Check for parsing problems
problems <- function(datapackage){
  
  dep <- datapackage$data$deployments
  obs <- datapackage$data$observations
  
  # check for locations with the same coordinates
  loc_dup_coords <- dep |> 
    group_by(locationID, latitude, longitude) |>
    distinct() |> 
    ungroup() |> 
    group_by(latitude, longitude) |> 
    summarize(n_locs = n()) |> 
    filter(n_locs > 1)
  
  # check for locations with same name
  loc_dup_names <- dep |> 
    group_by(locationName, latitude, longitude) |> 
    distinct() |> 
    ungroup() |> 
    group_by(locationName) |> 
    summarise(n_locs = n()) |> 
    filter(n_locs > 1)
  
  # check for locations with same locationID
  loc_dup_IDs <- dep |> 
    group_by(locationID, latitude, longitude) |> 
    distinct() |> 
    ungroup() |> 
    group_by(locationID) |> 
    summarise(n_locs = n()) |> 
    filter(n_locs > 1)
  
  # check for records outside the sampling period at the related deployment
  seq_out <- obs |> 
    left_join(dep |> select(deploymentID, locationID, locationName, deploymentStart, deploymentEnd)) |> 
    mutate(issues = ifelse(eventStart < deploymentStart | eventEnd > deploymentEnd, "Yes", "No")) |> 
    filter(issues == "Yes") |> 
    select(-issues)
  
  # check for multiple simultaneously-active deployments at the same location
  dep_overlap <- dep |> 
    arrange(locationID, locationName, deploymentStart) |> 
    group_by(locationID, locationName) |> 
    mutate(issues = ifelse(deploymentStart < lag(deploymentEnd), "Yes", "No")) |> 
    filter(issues == "Yes") |> 
    select(-issues)
  
  # build output object
  res <- list(loc_dup_coords, loc_dup_names, loc_dup_IDs, seq_out, dep_overlap)
  return(res)
}

ctdp_problems <- problems(ctdp)
if(all(lapply(ctdp_problems, nrow) == 0)) {
  print("No parsing problems found.")
} else {
  print("Parsing problems found:")
  # for (i in seq_along(ctdp_problems)) {
  #   if (nrow(ctdp_problems[[i]]) > 0) {
  #     print(paste(names(ctdp_problems)[i], "has parsing problems:"))
  #     print(ctdp_problems[[i]])
  #   }
  # }
  if (nrow(ctdp_problems[[1]]) > 0) {
    print(paste("There are", nrow(ctdp_problems[[1]]), "cases in which the same coordinates are associated with more than one location. Check the object ctdp_problems[[1]] for details."))
    print(ctdp_problems[[1]])
  }
  
  if (nrow(ctdp_problems[[2]]) > 0) {
    print(paste("There are", nrow(ctdp_problems[[2]]), "cases in which two or more locations have the same locationName. Check the object ctdp_problems[[2]] for details."))
    print(ctdp_problems[[2]])
  }
  
  if (nrow(ctdp_problems[[3]]) > 0) {
    print(paste("There are", nrow(ctdp_problems[[3]]), "cases in which two or more locations have the same locationID. . Check the object ctdp_problems[[3]] for details."))
    print(ctdp_problems[[3]])
  }
  
  if (nrow(ctdp_problems[[4]]) > 0) {
    print(paste("There are", nrow(ctdp_problems[[4]]), "observations dated outside the sampling period of the related deployment. Check the object ctdp_problems[[4]] for details."))
    print(ctdp_problems[[4]])
  }
  
  if (nrow(ctdp_problems[[5]]) > 0) {
    print(paste("There are", nrow(ctdp_problems[[5]]), "deployments with sampling periods that (at least partially) overlap deployment at the same location. Check the object ctdp_problems[[5]] for details."))
    print(ctdp_problems[[5]])
  }
} 
```

# 3. Filter data that is within the desired time period and export

The EOW survey period runs from 1st September to 31st October.
However, a 15-day buffer was set, such that all records from 15th August to 15th November are retained.
We will use the `filter_deployments()` and `filter_media()` functions to filter the dataset to this time period.
After that, we will truncate the deployment start (`deploymentStart`) and end (`deploymentEnd`) dates to the EOW survey period, in case that they are outside this period.
I.e. a deployment starting on 3th August will be truncated to 15th August, and a deployment ending on 20th November will be truncated to 15th November.
Finally, we will export the filtered dataset to separate folders for each year present in the data.

```{r filter and export}
# Specify the export directory
exdir <- normalizePath(tempdir(), winslash = "/", mustWork = FALSE)

# Loop over all years in the dataset
years <- year(ctdp$temporal$start):year(ctdp$temporal$end)
no_deps <- NULL

for (y in years) {
  eow_start <- ymd(paste0(y, "-08-15")) # EOW survey starts on 15th August
  eow_end <- ymd(paste0(y, "-11-15")) # EOW survey ends on 15th November
  
  # Filter deployments that overlap the EOW survey period 
  # (15th August - 15th November)
  out <- ctdp |> 
    filter_deployments(
      year(deploymentStart) == y,
      int_overlaps(
        interval(as.Date(deploymentStart), as.Date(deploymentEnd)),
        interval(eow_start, eow_end)
      )
    )
  
  # Discard media files outside the EOW survey period 
  # (this automatically filters observations as well)
  # but we will filter observations explicitly as well
  out <- out |> 
    filter_media(
      as.Date(timestamp) %within% interval(eow_start, eow_end)
    ) |> filter_observations(
      as.Date(eventStart) %within% interval(eow_start, eow_end),
      as.Date(eventEnd) %within% interval(eow_start, eow_end)
    )

  # Truncate the deploymentStart and deploymentEnd to the EOW survey period
  deployments(out) <- deployments(out) |>
    mutate(
      deploymentStart = pmax(deploymentStart, eow_start),
      deploymentEnd = pmin(deploymentEnd, eow_end)
    )
  
  if (camtrapdp:::version(out) == "1.0.1") {
    # Set version to 1.0, since there are no rs.gbif.org schemas for 1.0.1
    camtrapdp:::version(out) <- "1.0"
    out$directory <- exdir
    out$platform <- out$sources[[1]]
    out$name <- paste("EOW", proj, y, sep = "-")
  }
  
  # replace NA by an empty string in taxonomic data
  replace_na_recursive <- function(x) {
    if (is.null(x)) {
      return(NULL)
    } else if (is.atomic(x)) {
      # atomic vectors like character, numeric, etc.
      x[is.na(x)] <- ""
      return(x)
    } else if (is.list(x)) {
      return(lapply(x, replace_na_recursive))
    } else {
      return(x)  # for other types (e.g. functions, language objects)
    }
  }
  out$taxonomic <- replace_na_recursive(out$taxonomic)
  
  if (nrow(deployments(out)) == 0) {
    message("No deployments found for year ", y)
    no_deps <- c(no_deps, y)
    next
  }
  write_camtrapdp(out, file.path(exdir, paste("EOW", proj, y, sep = "-")))
}
```

# 4. Check if exported datasets can be imported again

Finally, we will also check if all dates and datetimes are within the EOW survey period. 
But this should not be necessary, since we already filtered the data to this period.

```{r check export}
for (y in years) {
  eow_start <- ymd(paste0(y, "-08-15")) # EOW survey starts on 15th August
  eow_end <- ymd(paste0(y, "-11-15")) # EOW survey ends on 15th November
  
  if (y %in% no_deps) {
    message("No deployments found for year ", y)
    next
  }
  
  message("Trying to import EOW dataset for year ", y)
  test <- camtrapdp::read_camtrapdp(file.path(exdir, paste("EOW", proj, y, sep = "-"), "datapackage.json"))
  
  # Check if all dates and datetimes are within the EOW survey period
  if(!is.null(test)) {
    check_datetimes <- function(out, eow_start, eow_end) {
      eow_start_time <- ymd_hms(paste(eow_start, "00:00:00"))
      eow_end_time   <- ymd_hms(paste(eow_end, "23:59:59"))
    
      ds <- deployments(out)
      ms <- media(out)
      obs <- observations(out)
    
      issues_found <- FALSE
      messages <- list()
    
      # Collect warnings instead of printing immediately
      idx <- which(ds$deploymentStart < eow_start)
      if (length(idx) > 0) {
        messages <- c(messages, sprintf("⚠️ Deployments starting before EOW start: rows [%s]", paste(idx, collapse = ", ")))
        issues_found <- TRUE
      }
    
      idx <- which(ds$deploymentEnd > eow_end)
      if (length(idx) > 0) {
        messages <- c(messages, sprintf("⚠️ Deployments ending after EOW end: rows [%s]", paste(idx, collapse = ", ")))
        issues_found <- TRUE
      }
    
      idx <- which(ms$timestamp < eow_start_time)
      if (length(idx) > 0) {
        messages <- c(messages, sprintf("⚠️ Media with timestamps before EOW start: rows [%s]", paste(idx, collapse = ", ")))
        issues_found <- TRUE
      }
    
      idx <- which(obs$eventStart > eow_end_time)
      if (length(idx) > 0) {
        messages <- c(messages, sprintf("⚠️ Observations with eventStart after EOW end: rows [%s]", paste(idx, collapse = ", ")))
        issues_found <- TRUE
      }
    
      idx <- which(obs$eventStart < eow_start_time)
      if (length(idx) > 0) {
        messages <- c(messages, sprintf("⚠️ Observations with eventStart before EOW start: rows [%s]", paste(idx, collapse = ", ")))
        issues_found <- TRUE
      }
    
      # Print results
      if (issues_found) {
        cat(paste(sapply(messages, function(m) paste0(m, "\n\n")), collapse = ""))
      } else {
        message("✅ All checks passed. No data issues detected.")
      }
    }
    
    check_datetimes(test, eow_start, eow_end)
  }
}
```
